---
title: "Supplementary Information: Figures and Tables"
header-includes:
   - \usepackage{caption}
   - \usepackage{multirow}
output: pdf_document
---

Beyond counts and averages: relating geodiversity to dimensions of biodiversity. Quentin D. Read, Phoebe L. Zarnetske, Sydne Record, Kyla M. Dahlin, Andrew O. Finley, John M. Grady, Martina L. Hobi, Sparkle L. Malone, Jennifer K. Costanza, Adam M. Wilson, Andrew M. Latimer, Keith D. Gaddis, Stephanie Pau, Scott V. Ollinger

\newcommand{\fpmap}{/Users/qread/google_drive/NASABiodiversityWG/Figures/multivariate_maps_figs/maps_for_supplement}

# Supplementary Tables

**Table S1**: Overview of biodiversity metrics by level and by dimension.

\begin{table}[h]
\begin{tabular}{m{1cm}m{2cm}p{4cm}p{4cm}p{4cm}}
       &                                                            & \multicolumn{3}{c}{\textit{\textbf{Diversity dimension}}} \\
       &                                                            & \textbf{Taxonomic}  & \textbf{Phylogenetic}  & \textbf{Functional} 
\\[4pt]
\multirow{3}{*}{\rotatebox{90}{\textit{\textbf{Diversity level}}}} 
      & \begin{tabular}[c]{@{}l@{}}\\\textbf{Alpha}\\\textbf{(local)}\end{tabular}  
      & Mean richness of local communities within a 50-km radius around the focal plot
      & Mean pairwise phylogenetic distance score of local communities in the radius, relative to null expectation
      & Mean pairwise functional distance score of local communities in the radius, relative to null expectation
\\[4pt]
      & \begin{tabular}[c]{@{}l@{}}\\\textbf{Beta}\\\textbf{(turnover)}\end{tabular} 
      & Mean pairwise dissimilarity of local communities in the radius, logit-transformed
      &  Mean pairwise phylogenetic distance among communities in the radius, relative to null expectation
      & Mean pairwise functional distance among communities in the radius, relative to null expectation
\\[4pt]
      & \begin{tabular}[c]{@{}l@{}}\\\textbf{Gamma}\\\textbf{(regional)}\end{tabular} 
      &  Richness of all communities in the radius, pooled to a single community
      &  Mean pairwise phylogenetic distance of all communities in the radius, pooled to a single community, relative to null expectation
      &  Mean pairwise functional distance of all communities in the radius, pooled to a single community, relative to null expectation
      
\end{tabular}
\end{table}

\newpage

**Table S2**: This table shows model fit statistics for each response variable for each model. The RMSE is the root mean squared error of the model fit with all the data. The CV RMSE is the root mean squared error of the holdout data points in spatially blocked cross-validation. Both RMSEs are divided by the range of the data to produce a relative RMSE value that is comparable among models. Lower RMSE values indicate better model performance. The $R^2$ is calculated using the method of Gelman *et al.*. For all fit statistics, mean value as well as upper and lower bounds of 95% credible intervals are shown. 

```{r, echo = FALSE, message = FALSE}
library(knitr)
library(kableExtra)
fit_stats <- read.csv('~/google_drive/NASABiodiversityWG/FlavorsOfDiversityPaper/supptable_fitstats_withrelativeRMSEs.csv')
fit_stats$model <- gsub('+', ' + ', fit_stats$model, fixed = TRUE)
kable(fit_stats, 'latex', escape = FALSE, longtable = TRUE, booktabs = TRUE,
             linesep = c('', '', '', '\\addlinespace'),
             row.names = FALSE,
             col.names = c('taxon','biodiversity level', 'biodiversity dimension', 'model', 'RMSE', 'CV RMSE', '$R^2$')) %>%
  kable_styling(latex_options = 'repeat_header') %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(2, width = '6em') %>%
  column_spec(3, width = '6em') %>%
  column_spec(4, width = '9em')
```

\newpage

**Table S3**: This table shows the Widely Available Information Criterion (WAIC) for each model, along with the standard error of the WAIC. Lower WAIC values indicate a more parsimonious fit.

```{r, echo = FALSE, message = FALSE}
waic_table <- read.csv('~/google_drive/NASABiodiversityWG/FlavorsOfDiversityPaper/supptable_WAICs.csv')
waic_table$model <- gsub('+', ' + ', waic_table$model, fixed = TRUE)
kable(waic_table, 'latex', escape = FALSE, longtable = FALSE, booktabs = TRUE,
             linesep = c('', '', '', '\\addlinespace'),
             row.names = FALSE,
             col.names = c('taxon','biodiversity level', 'model', 'WAIC', 'standard error of WAIC')) %>%
  row_spec(0, bold = TRUE) 
```

\newpage

# Supplementary Figures

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{/Users/qread/google_drive/NASABiodiversityWG/Figures/multivariate_maps_figs/07may2019/both_lolormse_allmodels_2wayfacet.png}}
\caption*{Figure S1: Model performance assessed with five-fold root mean squared error}
\end{figure}

**Figure S1**: This figure shows the relative root mean squared errors obtained from spatially blocked leave-one-location-out cross-validation, by comparing the observed value with the out-of-sample predicted value on the data points omitted in each fold. A lower value represents better performance of the model predicting biodiversity. Error bars are 95% credible intervals. Here, we not only compare taxa but also the four types of model: space only (null), space + climate, space + geodiversity, and space + climate + geodiversity (full model). Due to the fact that each cross-validation fold excludes an entire region, the null model including only the spatial random effect tends to predict the held out values as well or better than the models including climate and geodiversity predictors. However, models including geodiversity predictors tend to perform as well or better than the models including climate predictors, especially for phylogenetic and functional diversity.

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{/Users/qread/google_drive/NASABiodiversityWG/Figures/multivariate_maps_figs/07may2019/both_multivariate_coefvar_sideways.png}}
\caption*{Figure S2: Spatial variation in geodiversity-biodiversity relationships}
\end{figure}

**Figure S2**: The spatial variability of the relationship between each predictor and each response, by taxon. The spatial variability is measured by the standard deviation of each coefficient value among ecoregions. Error bars represent the 95% credible interval of the standard deviation parameter. Higher values represent relationships that vary more spatially.

\newpage

### Supplementary Figures S3-S14

The following figures are maps of the coefficient on each climate or geodiversity predictor from the full multivariate models predicting alpha, beta, and gamma diversity for birds and for trees. Each figure shows the coefficients from the alpha, beta, and gamma models for each of the three biodiversity dimensions (taxonomic, phylogenetic, and functional) for one taxonomic group. Figures 3 through 8 show the coefficients for birds separately for each predictor, and figures 9 through 14 show the coefficients for trees.

In figures S3a, S4a, etc. below, the color of each TNC ecoregion is the region-level coefficient estimate from the model (USA-wide fixed-effect slope + random-effect slope from that region). The colors are scaled so that darker red ecoregions have more positive coefficients, and darker blue ecoregions have negative coefficients. White ecoregions have coefficients approximating zero, and dark gray ecoregions have coefficients that could not be estimated. 

In figures S3b, S4b, etc., below, the ecoregions are shaded based on whether the 95% credible interval around the coefficient overlaps zero. Red shaded regions have entirely positive nonzero 95% credible intervals, while blue shaded regions have entirely negative nonzero 95% credible intervals. Dark gray shaded regions have 95% credible intervals that contain zero.

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_coef_temperature_mean.png}}
\caption*{Figure S3a: Coefficients of mean annual temperature on bird diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_signif_temperature_mean.png}}
\caption*{Figure S3b: Coefficients of mean annual temperature on bird diversity, nonzero regions shown}
\end{figure}

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_coef_precip_mean.png}}
\caption*{Figure S4a: Coefficients of mean annual precipitation on bird diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_signif_precip_mean.png}}
\caption*{Figure S4b: Coefficients of mean annual precipitation on bird diversity, nonzero regions shown}
\end{figure}

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_coef_elevation_diversity.png}}
\caption*{Figure S5a: Coefficients of elevation variability on bird diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_signif_elevation_diversity.png}}
\caption*{Figure S5b: Coefficients of elevation variability on bird diversity, nonzero regions shown}
\end{figure}

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_coef_gpp_sd.png}}
\caption*{Figure S6a: Coefficients of GPP variability on bird diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_signif_gpp_sd.png}}
\caption*{Figure S6b: Coefficients of GPP variability on bird diversity, nonzero regions shown}
\end{figure}

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_coef_geol_age_diversity.png}}
\caption*{Figure S7a: Coefficients of geological age diversity on bird diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_signif_geol_age_diversity.png}}
\caption*{Figure S7b: Coefficients of geological age diversity on bird diversity, nonzero regions shown}
\end{figure}

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_coef_soil_diversity.png}}
\caption*{Figure S8a: Coefficients of soil type diversity on bird diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/bbs_signif_soil_diversity.png}}
\caption*{Figure S8b: Coefficients of soil type diversity on bird diversity, nonzero regions shown}
\end{figure}

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_coef_temperature_mean.png}}
\caption*{Figure S9a: Coefficients of mean annual temperature on tree diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_signif_temperature_mean.png}}
\caption*{Figure S9b: Coefficients of mean annual temperature on tree diversity, nonzero regions shown}
\end{figure}

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_coef_precip_mean.png}}
\caption*{Figure S10a: Coefficients of mean annual precipitation on tree diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_signif_precip_mean.png}}
\caption*{Figure S10b: Coefficients of mean annual precipitation on tree diversity, nonzero regions shown}
\end{figure}

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_coef_elevation_diversity.png}}
\caption*{Figure S11a: Coefficients of elevation variability on tree diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_signif_elevation_diversity.png}}
\caption*{Figure S11b: Coefficients of elevation variability on tree diversity, nonzero regions shown}
\end{figure}

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_coef_gpp_sd.png}}
\caption*{Figure S12a: Coefficients of GPP variability on tree diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_signif_gpp_sd.png}}
\caption*{Figure S12b: Coefficients of GPP variability on tree diversity, nonzero regions shown}
\end{figure}

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_coef_geol_age_diversity.png}}
\caption*{Figure S13a: Coefficients of geological age diversity on tree diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_signif_geol_age_diversity.png}}
\caption*{Figure S13b: Coefficients of geological age diversity on tree diversity, nonzero regions shown}
\end{figure}

\newpage

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_coef_soil_diversity.png}}
\caption*{Figure S14a: Coefficients of soil type diversity on tree diversity, values shown}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics[width=0.75\textwidth]{\fpmap/fia_signif_soil_diversity.png}}
\caption*{Figure S14b: Coefficients of soil type diversity on tree diversity, nonzero regions shown}
\end{figure}

